Who Gets What: How Population and Income Shape Toronto’s Budget Decisions

### 2024-12-03 paper/dump.qmd
Added New Test Scripts
- Added Simulation
- Added Simulation Tests
- Added Analysis Data Tests
- Added Sketches

## Park and Recreation Facility Projects {#sec-parks}

The Park and Recreation Facility Projects dataset [@toronto_parks], published by the Parks, Forestry and Recreation Division and last updated on July 22, 2024, contains information about park and recreation facility projects in Toronto. This dataset includes details about the project name, location, ward number, and project status. Park and recreation facilities are essential public amenities that contribute to the quality of life and well-being of residents, providing spaces for recreation, leisure, and community engagement.

We choose this dataset to evaluate the how budget allocations are related to the number of park and recreation facility projects in each ward. The number of park projects can serve as a proxy for community development and public service enhancement, reflecting the need for investments in public amenities and infrastructure. By analyzing the relationship between park projects and budget allocations, we can gain insights into how public service enhancements influence the distribution of resources across the city. This will also help us understand how the city's budget decisions align with community development trends and public service needs in different wards, and determine if there are any disparities in resource allocation based on public service enhancements to certain wards.

```{r}
parks_data <- here::here("data/01-raw_data/Park and Recreation Facility Projects - 4326.csv")
parks <- read_csv(parks_data)

# create a graph of the number of parks by ward_number
parks |>
  count(ward_number) |>
  ggplot(aes(x = ward_number, y = n)) +
  geom_col() +
  labs(x = "Ward number", y = "Number of parks") +
  theme_minimal()
```

library(tidyverse)
library(palmerpenguins)
```

@@

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

@@

library(boot)
library(janitor)
library(modelsummary)
library(rstanarm)
library(tidyverse)
library(collapse)
library(tinytable)
library(marginaleffects)
```

```{r}
# combine all the years data for each ward and for each category
model_data <- analysis_budget_data |>
  group_by(ward_id, category, ward) |>
  summarize(total_10_year = sum(total_10_year)) |>
  pivot_wider(names_from = category, values_from = total_10_year)

# combine the model data with the ward profile data
model_data <- model_data |>
  left_join(analysis_ward_data, by = "ward_id") |>
  select(ward_id, ward.x, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
  rename(ward = ward.x) |>
  left_join(analysis_building_permits, by = "ward_id") |>
  rename(growth_related = `Growth Related`, state_of_good_repair = `State of Good Repair`, service_improvement = `Service Improvement and Enhancement`)

```{r}
# Install and load rstanarm if not installed
# install.packages("rstanarm")
# Load required libraries
library(rstanarm)
library(bayesplot)

# Prepare the data (assuming you've read the CSV)
data <- read.csv(here::here("data/02-analysis_data/model_data.csv"))

write.csv(data, here::here("data/02-analysis_data/model_data.csv"))

# Basic Poisson Model
permits_model <- stan_glm(
  building_permits ~ ward_id,
  data = model_data,
  family = poisson(link = "log"),s
  prior = normal(0, 2.5),  # Weakly informative prior
  prior_intercept = normal(0, 10),
  chains = 4,
  iter = 2000,
  seed = 123
)

# Model summary
summary(permits_model)

# Diagnostic plots
plot(permits_model)

# Posterior predictive checks
pp_check(permits_model)

# Credible intervals
posterior_interval(permits_model)

# Visualize posterior distributions
mcmc_areas(permits_model)

# Predictive performance
# Compare observed vs. predicted
predicted_permits <- posterior_predict(permits_model)

```{r}
library(rstanarm)
library(bayesplot)

# Fit the Negative Binomial model
# building_permits is the target variable
# population, average_household_income, growth_related, state_of_good_repair, and service_improvement are predictors
model <- stan_glmer(
  growth_related ~ population + average_household_income + (1 | ward_id),
  data = data,
  family = neg_binomial_2(link = "log"),
  chains = 4,
  iter = 2000,
  seed = 1234
)

# Summarize the model
summary(model)

@@

# Diagnostic plots
plot(model)

@@

# Predictive performance
# Compare observed vs. predicted
predicted_permits <- posterior_predict(model)
```

```{r}
# model summary

>>>>

@@

@@

library(boot)
library(janitor)
library(modelsummary)
library(rstanarm)
library(tidyverse)
library(collapse)
library(tinytable)
library(marginaleffects)
library(rstanarm)
library(bayesplot)
```

```{r}
analysis_budget_data <- read.csv(here::here("data/02-analysis_data/analysis_budget_data.csv"))
analysis_building_permits <- read.csv(here::here("data/02-analysis_data/analysis_building_permits.csv"))
analysis_ward_data <- read.csv(here::here("data/02-analysis_data/analysis_ward_data.csv"))

# combine all the years data for each ward and for each category
model_data <- analysis_budget_data |>
  group_by(ward_id, category, ward) |>
  summarise(avg_total_10_year = mean(total_10_year)) |>
  pivot_wider(names_from = category, values_from = avg_total_10_year) |>
  # sum the total 10 year budget for each ward
  group_by(ward_id, ward) |>
  summarise(total_budget = sum(`Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`))

# combine the model data with the ward profile data
model_data <- model_data |>
  left_join(analysis_ward_data, by = "ward_id") |>
  select(ward_id, ward.x, total_budget, average_household_income, population) |>
  rename(ward = ward.x) |>
  # left join with the building permits data
  left_join(analysis_building_permits, by = "ward_id") |>
  select(ward_id, ward, total_budget, average_household_income, population, total_building_permits)
```

```{r}

# convert the ward_id to a factor
model_data$ward_id <- as.factor(model_data$ward_id)

model <- stan_glm(
  formula = total_building_permits ~ total_budget + average_household_income + population + ward_id,
  data = model_data,
  family = gaussian(link = "identity"),
  prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
  prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE)
)

# Summarize model output
summary(model)

@@

# Plot the coefficient estimates
plot(model)

@@

# Predictive performance
# Compare observed vs. predicted
predicted_permits <- posterior_predict(model)

# plot diagnostics
plot(model, "trace")
plot(model, "rhat")

modelsummary(
  list(
    "Model" = model
  ),
)
```

```{r}

### 2024-12-03 paper/paper.qmd
Added New Test Scripts
- Added Simulation
- Added Simulation Tests
- Added Analysis Data Tests
- Added Sketches

## Model {#sec-model}

The purpose of this paper is to investigate the relationship between population density, average household income, and budget allocations across Toronto's 25 wards, and their potential impact on the economic well-being of each ward, as represented by the `total_building_permits` issued. To achieve this, we developed a causal model that examines how population and income influence the allocation of funds in key budget categories: `Growth Related`, `State of Good Repair`, and `Service Improvement and Enhancement`.

We employed multilevel modeling to estimate the effect of these variables on the number of building permits issued in each ward. Here we briefly describe a Bayesian Poisson regression model, which incorporates varying intercepts by grouping the effect by ward account. This approach allows us to capture the nuanced effects of demographic and economic factors on infrastructure development and resource distribution. Background details and diagnostics are included in [Appendix @sec-model-details].

### Model set-up

The model was specified as follows:

\begin{align*}
\text{Building Permits}_{i} & \sim \text{Poisson}(\lambda_{i}) \\
\log(\lambda_{i}) & = \beta_{0} + \beta_{1} \times \text{Population}_{i} + \beta_{2} \times \text{Income}_{i} + \beta_{3} \times \text{Budget}_{i} + \alpha_{\text{ward}[i]} \\
\beta_{0} & \sim \text{Normal}(0, 2.5) \\
\beta_{1} & \sim \text{Normal}(0, 2.5) \\
\beta_{2} & \sim \text{Normal}(0, 2.5) \\
\beta_{3} & \sim \text{Normal}(0, 2.5) \\
\alpha_{\text{ward}} & \sim \text{Normal}(0, \sigma_{\text{ward}})
\end{align*}

-   $\text{Building Permits}_i$ represents the number of building permits issued in ward `i`,
-   $\lambda_i$ is the expected count of permits for ward $i$.
-   $\log(\lambda_i)$ is the linear predictor, which includes:
    -   $\beta_0$: Intercept
    -   $\beta_1$: Effect of `population`
    -   $\beta_2$: Effect of `average_household_income`
    -   $\beta_3$: Effect of `budget_allocation`
    -   $\alpha_{\text{ward}[i]}$: Random effect for each ward (captures unobserved heterogeneity between wards).
-   Priors for $\beta_0$, $\beta_1$, $\beta_2$, and $\beta_3$ are Normal(0, 2.5), allowing for flexibility.

The model was fitted using the `stan_glmer` function from the `rstanarm` package [@citerstanarm] in R [@citeR]. The prior distributions for the coefficients were set to normal distributions with a mean of 0 and a standard deviation of 2.5. The model also included a prior for the intercept, which was set to a normal distribution with a mean of 0 and a standard deviation of 2.5. The varying intercepts for each ward were modeled using a normal distribution with a mean of 0 and a standard deviation of $\sigma_{\text{ward}}$.

### Model justification

## Results {#sec-results}

### Relationship between Population, Average Household Income, and Budget Allocations

@@

average_budget |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
                      "State of Good Repair", "Service Improvement and Enhancement"))
```
## Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects

@@

combined_data <- analysis_ward_data |>
  left_join(average_budget, by = "ward_id") |>
  rename(ward = ward.x) |>
  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`)

@@

## Relationship between Average Household Income & Population with Budget Allocation for State of Good Repair Projects

@@

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
```

@@

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
```

@@

@@

# Plot with labels for outliers
ggplot(service_improvement, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    title = "Total Budget Over Years",
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| warning: false
#| message: false
# select only growth related, state of good repair and service improvement and enhancement
growth_related <- analysis_budget_data |>
  filter(category == "Growth Related") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Plot with labels for outliers
ggplot(growth_related, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    title = "Total Budget Over Years",
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| warning: false
#| message: false
# select only growth related, state of good repair and service improvement and enhancement
state_of_good_repair <- analysis_budget_data |>
  filter(category == "State of Good Repair") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Plot with labels for outliers
ggplot(state_of_good_repair, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    title = "Total Budget Over Years",
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

>>>>

```{r}
#| echo: false
#| warning: false
#| message: false

# select only growth related
growth_related <- analysis_budget_data |>
  filter(category == "Growth Related") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Plot with labels for outliers
plot1 <- ggplot(growth_related, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| warning: false
#| message: false

# select only growth related, state of good repair and service improvement and enhancement
state_of_good_repair <- analysis_budget_data |>
  filter(category == "State of Good Repair") |>
  mutate(total_10_year = total_10_year / 1000) |>
  # remove negative values
  filter(total_10_year > 0) |>
 rename(total_budget = total_10_year)

# Plot with labels for outliers
plot2 <- ggplot(state_of_good_repair, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

# Plot with labels for outliers
plot3 <- ggplot(service_improvement, aes(x = factor(year), y = total_budget)) +
  geom_boxplot(fill = "skyblue", color = "darkblue") +
  geom_jitter(aes(color = as.factor(ward_id)), width = 0.2) +
  geom_text(
    data = outliers, 
    aes(label = ward), 
    nudge_y = 15, 
    check_overlap = TRUE
  ) +  # Label only outliers
  labs(
    x = "Year",
    y = "Total Budget (in thousands)"
  ) +
  theme_minimal() +
  # remove legend
  theme(legend.position = "none")
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-budget-trends
#| fig.cap: "Trends in budget allocations (2021-2024) by category"
#| fig-subcap: ["Growth Related", "State of Good Repair", "Service Improvement and Enhancement"]
#| fig-pos: "h"
#| layout-ncol: 2

plot1
plot2
plot3
```

## Relationship between Population, Average Household Income, and Budget Allocations

@@

average_budget |>
  head() |>
  kable(col.names = c("Ward ID", "Ward Name", "Growth Related", 
                      "State of Good Repair", "Service Improvement and Enhancement"))
```
### Relationship between Average Household Income & Population with Budget Allocation for Growth Related Projects and State of Good Repair Projects

@@

combined_data <- analysis_ward_data |>
  left_join(average_budget, by = "ward_id") |>
  left_join(analysis_building_permits, by = "ward_id") |>
  rename(ward = ward.x) |>
  select(ward_id, ward, population, average_household_income, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`, total_building_permits)

@@

@@

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
```
[@fig-growth-related-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Growth-Related projects. The linear regression line for average household income remains relatively flat, hovering around \$100,000. This suggests that budget allocations for Growth-Related projects are not significantly influenced by income levels, indicating a more uniform distribution of funds across wards regardless of household income. A similar lack of correlation is observed in [@fig-state-of-good-repair-budget], where budget allocations for State of Good Repair projects also show minimal dependence on average household income.

In contrast, the linear regression line for population shows a positive correlation with budget allocation for Growth-Related projects. Wards with higher population densities—such as Ward 10 (Spadina-Fort York), Ward 13 (Toronto Centre), Ward 5 (York South-Weston), and Ward 2 (Etobicoke North)—receive a larger share of the Growth-Related budget. This indicates that population density plays a critical role in determining funding levels for these projects, with more populous wards benefiting from increased allocations. A similar trend is observed in the budget distribution for State of Good Repair projects, where wards with higher population densities receive a proportionally larger share of the funds.

@@

# Combine both plots side by side
grid.arrange(p1, p2, ncol = 2)
```
[@fig-service-improvement-budget] illustrates the relationship between average household income, population, and the total 10-year budget allocation for Service Improvement and Enhancement projects. These projects  aim to enhance the quality of services provided to residents and improve the overall infrastructure of the city [@toronto_improvements].

Unlike Growth-Related and State of Good Repair projects, the linear regression lines for average household income and population shows a positive correlation with budget allocation for Service Improvement and Enhancement projects. Wards with higher average household incomes, such as Ward 11 (University-Rosedale) and Ward 13 (Toronto Centre), receive a larger share of the budget allocation for these projects. This suggests that higher-income areas are prioritized for service improvements and infrastructure enhancements, reflecting a targeted approach to resource allocation based on income levels.

Other wards like Ward 3 (Etobicoke-Lakeshore) and Ward 10 (Spadina-Fort York), which have high population densities, also received significant budget allocations, suggesting that these areas are priorities for service improvements and infrastructure enhancement.

Despite these outliers, the majority of wards received similar budget allocations for Service Improvement and Enhancement projects, underscoring a city-wide strategy to enhance services uniformly for all residents.

# Model {#sec-model}

The purpose of this paper is to investigate the relationship between population density, average household income, and budget allocations across Toronto's 25 wards, and their potential impact on the economic well-being of each ward, as represented by the `total_building_permits` issued. To achieve this, we will first plot the data to visualize the relationships between these variables and the number of building permits issued in each ward independently. Then we will fit a Bayesian regression model with varying intercepts to estimate the effect of these variables on the number of building permits issued in each ward.

## Model set-up

The particular model we used is Bayesian multiple linear regression model with varying intercepts. The model includes the following variables:

\begin{align*} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 + \beta_1 x_{\text{population}} + \beta_2 x_{\text{income}} + \beta_3 x_{\text{budget}} + \gamma_i \\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
\gamma_i &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align*}

- $\mu_i$ is the expected number of building permits issued in ward $i$.
- $\beta_0$ is the intercept, representing the expected number of building permits issued in a ward with average values for all other variables.
- $\beta_1$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the population density of the ward.
- $\beta_2$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the average household income of the ward.
- $\beta_3$ is the coefficient for the predicted change in the number of building permits issued in a ward given a one unit increase in the total budget allocation across three categories: Growth Related, State of Good Repair, and Service Improvement and Enhancement.

The model also includes a varying intercept $\gamma_i$ for each ward, accounting for the unobserved heterogeneity between wards. The varying intercepts allow the model to capture the unique characteristics of each ward that may influence the number of building permits issued.

The model was fitted using the `stan_glmer` function from the `rstanarm` package [@citerstanarm] in R [@citeR]. The prior distributions for the coefficients were set to normal distributions with a mean of 0 and a standard deviation of 2.5. The model also included a prior for the intercept, which was set to a normal distribution with a mean of 0 and a standard deviation of 2.5. The standard deviation of the likelihood was set to an exponential distribution with a rate parameter of 1.

## Model justification

The above factors were choosen as they are known to influence the number of building permits issued in each ward. Population density and average household income are key demographic and economic indicators that can impact construction activity and development. Budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects are critical determinants of infrastructure development and resource distribution. By incorporating these variables into the model, we can assess their individual and combined effects on the number of building permits issued in each ward.

A priori, we expect that population density and average household income will have a positive effect on the number of building permits issued, reflecting the demand for construction and development in densely populated and affluent areas. Similarly, we anticipate that budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects will positively influence the number of building permits, reflecting the city's investment in infrastructure development and maintenance.

# Results {#sec-results}

## Relationship between Building Permits and Population, Income, and Budget Allocations

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-building-permits-relationship
#| fig.cap: "Relationship between Variables and Total Building Permits"
#| fig-subcap: ["Population", "Average Household Income", "Total Budget in Three Categories"]
#| fig-pos: "h"
#| layout-ncol: 2

results_data <-
  analysis_budget_data |>
  group_by(ward_id, ward, category) |>
  summarise(total_10_year = mean(total_10_year)) |>
  pivot_wider(names_from = category, values_from = total_10_year) |>
  select(ward_id, ward, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
  left_join(analysis_building_permits, by = "ward_id") |>
  left_join(analysis_ward_data, by = "ward_id") |>
  rename(ward = ward.x) |>
  select(ward_id, ward, population, average_household_income, total_building_permits, `Growth Related`, `State of Good Repair`, `Service Improvement and Enhancement`) |>
  # add a new column for total budget across three categories
  mutate(total_budget = `Growth Related` + `State of Good Repair` + `Service Improvement and Enhancement`) |>
  select(-`Growth Related`, -`State of Good Repair`, -`Service Improvement and Enhancement`)

# Create multiple scatter plots
plot1 <- ggplot(results_data, aes(x = population, y = total_building_permits, label = ward_id)) +
    geom_point(color = "blue") +
    geom_text_repel(hjust = -0.5, vjust = 0.5) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(x = "Population", y = "Total Building Permits")

plot2 <- ggplot(results_data, aes(x = average_household_income, y = total_building_permits, label = ward_id)) +
    geom_point(color = "blue") +
    geom_text_repel(hjust = -0.5, vjust = 0.5) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(x = "Average Household Income", y = "Total Building Permits")

plot3 <-
  ggplot(results_data, aes(x = total_budget, y = total_building_permits, label = ward_id)) +
    geom_point(color = "blue") +
    geom_text_repel(hjust = -0.5, vjust = 0.5) +
    geom_smooth(method = "lm", color = "red", se = TRUE) +
    labs(
      x = "Total Budget in Three Categories (Sum)",
      y = "Total Building Permits",
    )

plot1
plot2
plot3
```

We now explore the results of independent comparisons between the outcome variable (total number of building permits in Toronto's 25 wards) and the predictor variables (population density, average household income, and budget allocations for Growth-Related, State of Good Repair, and Service Improvement and Enhancement projects). As shown in [@fig-building-permits-relationship], most predictor variables exhibit relatively flat linear regression lines, indicating a weak relationship with the total number of building permits.

For example, the relationship between population density and building permits is notably flat, suggesting that population density has minimal influence on construction activity and development across the city’s wards—an unexpected result. A similar trend is observed for Growth-Related and State of Good Repair budget allocations, where flat regression lines suggest a weak relationship. This appears contradictory to earlier findings, where higher population density corresponded to increased budget allocations for these categories, which are typically associated with infrastructure development and maintenance. The same weak correlation is seen for Service Improvement and Enhancement budget allocations, as indicated by the flat regression line.

In contrast, the relationship between average household income and building permits shows a strong positive correlation. Higher-income areas tend to experience more construction activity and development. For instance, Ward 15 (Don Valley West) has the highest number of building permits, aligning with its high average household income. This suggests that higher-income areas are more likely to see increased construction and development, highlighting a clear link between income levels and building permits.

## Model Results {#sec-model-results}

@@

## Average Household Income

# Weaknesses and Limitations {#sec-limitations}

There are number of limitations that neeeds to be addressed in reagrds to the analysis. As we discussed in section @sec-data, the data used in this analysis is based on publicly available datasets from the City of Toronto. While these datasets provide valuable insights into the city's demographics, budget allocations, and building permits, they may not capture the full complexity of the city's infrastructure development and resource distribution. For example, the building permits dataset only includes active permits, which may not fully represent the scope of construction activity in Toronto. Additionally, the budget allocations dataset may not capture all infrastructure projects and expenditures, leading to potential gaps in the analysis.

Additionally, as shown in @sec-appendix, that there are 45 unique statuses assigned to each submitted permit. For this analysis, we just focused on one subset of these statuses. This may not fully capture the complexity of the building permit process and the various stages of construction and development in Toronto. Future analyses could explore additional permit statuses to gain a more comprehensive understanding of construction activity in the city.

One limitation of this study is the lack of detailed data on specific projects within each budget category, which could provide further insights into why certain wards receive more or less funding. Additionally, while this analysis focused on population density and average household income, other factors like infrastructure needs, community priorities, and political considerations likely play a role in budget decisions. Future research should explore these variables to build a more complete picture of how municipal budgets are allocated.

A potential limitation is regard ot the model. We assumed that the relationship between population density, average household income, and budget allocations is linear, which may not fully capture the complexity of these relationships. We also assumed independence of each ward, which may not hold true in practice due to shared resources, infrastructure projects, and other factors that could influence building permit issuance. Future research could explore more sophisticated modeling techniques to account for these dependencies and non-linear relationships.

# Future Work {#sec-future-work}

This analysis provides a foundation for future research on infrastructure development, resource distribution, and economic well-being in Toronto. Future studies could explore additional factors that influence budget allocations and building permit issuance, such as infrastructure needs, community priorities, and political considerations. By incorporating these variables into the analysis, researchers can gain a more comprehensive understanding of how municipal budgets are allocated and how construction activity is distributed across the city's wards.

In terms of policy recommendations, the city should adopt equity-based budgeting to ensure underfunded, high-density wards receive adequate resources, particularly for essential services like health and safety. Toronto could also implement a baseline funding model for high-need areas and enhance public engagement in the budget process, ensuring that residents of under served wards have a voice in resource allocation. These actions would promote more transparent, balanced, and equitable urban governance.

@@
